\begin{spacing}{1}
\begin{thebibliography}{100}
\addcontentsline{toc}{chapter}{References}


\bibitem{AIPoweredMaterialScience}
Bai X, Zhang X. Artificial Intelligence-Powered Materials Science. Nanomicro Lett. 2025;17(1):135. Published 2025 Feb 6. doi:10.1007/s40820-024-01634-8
\url{https://pmc.ncbi.nlm.nih.gov/articles/PMC11803041/}

\bibitem{weininger1988smiles}
Weininger, David. “SMILES, a chemical language and information system. 1. Introduction to methodology and encoding rules.” J. Chem. Inf. Comput. Sci. 28 (1988): 31-36.

\bibitem{lin2019bigsmiles}
Lin TS, Coley CW, Mochigase H, et al. BigSMILES: A Structurally-Based Line Notation for Describing Macromolecules. ACS Cent Sci. 2019;5(9):1523-1531. doi:10.1021/acscentsci.9b00476

\bibitem{rdkit}
RDKit: Open-source cheminformatics. https://www.rdkit.org

\bibitem{MDPolymerMoleculeInteraction}
Aulifa, D. L., Al Shofwan, A. A., Megantara, S., Fakih, T. M., \& Budiman, A. (2024). Elucidation of Molecular Interactions Between Drug-Polymer in Amorphous Solid Dispersion by a Computational Approach Using Molecular Dynamics Simulations. Advances and applications in bioinformatics and chemistry : AABC, 17, 1–19. https://doi.org/10.2147/AABC.S441628

\bibitem{DFTPolymerStructure}
Wei, Z.; Ning, N.; Zhang, L.; Tian, M.; Mi, J. Density Functional Theory of Polymer Structure and Conformations. Polymers 2016, 8, 121. https://doi.org/10.3390/polym8040121 

\bibitem{polymetrix}
Kunchapu, S., Jablonka, K.M. PolyMetriX: an ecosystem for digital polymer chemistry. npj Comput Mater 11, 312 (2025). https://doi.org/10.1038/s41524-025-01823-y

\bibitem{kingma2014adam}
Kingma, Diederik P. and Jimmy Ba. “Adam: A Method for Stochastic Optimization.” CoRR abs/1412.6980 (2014): n. pag.

\bibitem{vaswani2017attention}
Vaswani, Ashish, Noam Shazeer, Niki Parmar, Jakob Uszkoreit, Llion Jones, Aidan N. Gomez, Lukasz Kaiser, and Illia Polosukhin (2017), “Attention is All you Need,” Neural Information Processing Systems, 30, 5998–6008.

\bibitem{radford2018improving}
Radford, A., \& Narasimhan, K. (2018). Improving Language Understanding by Generative Pre-Training.






\end{thebibliography}
\end{spacing}


